---
title: "Homework 02"
author: "Kimberly Sharpe"
output:
  html_document:
    keep_md: yes
    theme: paper
  output:
    html_document:
      df_print: paged
  pdf_document: default
---

<!--- allow for errors while knitting-->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

<!--- load packages needed for the assignment ---> 

```{r load packages, echo=FALSE, message=FALSE}
library(gapminder)
library(tibble)
library(tidyverse)
library(ggplot2)
library(scales)
```
 
# Exercise 1: Basic dplyr 

#### 1.1 subsetting gapminder to three countries in the 1970s
```{r choose 3 countries in 1970s}
gapminder %>% 
  filter(year > 1969 & year < 1980, 
        country == "Afghanistan" |
        country == "Canada" | 
        country == "Denmark")
```


#### 1.2 using a pipe operator to select country & gdpPercap from filtered dataset
GDP per capita in the 1970s 
```{r}
gapminder %>% 
  filter(year > 1969 & year < 1980, 
        country == "Afghanistan" |
        country == "Canada" | 
        country == "Denmark") %>% 
  select(country, gdpPercap, year)
  
```

#### 1.3 filtering gapminder to all entries that have experienced a drop in life expectancy
```{r}
gapminder %>%
  arrange(year) %>% 
  group_by(country) %>% 
  mutate(diff_LifeExp = lifeExp - lag(lifeExp)) %>% 
  filter(diff_LifeExp < 0) %>% 
  arrange(diff_LifeExp)
```



#### 1.4 showing max GDP per capita experienced by each country 
```{r}
gapminder %>% 
  select(country, year, gdpPercap) %>% 
  group_by(country) %>% 
  filter(gdpPercap == max(gdpPercap)) 
```


#### 1.5 producing a scatterplot of Canada's life expectancy vs GDP 
```{r}
gapminder %>% 
  filter(country == "Canada") %>% 
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point(color = "green") +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) +
  theme_bw() +
  ylab("Life Expectancy")
```


# Exercise 2: Explore individual variables with dplyr

**Choose one categorial and one quantitative variable:**
Categorical variable: continent
Quantative variable: gdpPercap

## What are the possible values of each variable?

#### Continent: 

*How many continents are in the dataset?*
```{r}
gapminder %>% 
  select(continent) %>% 
  summarize(n_unique = n_distinct(continent))

```

*What continents are included in the dataset?*
```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarize(n_unique = n_distinct(continent))
```

*How many countries are in each continent?*
```{r}
gapminder %>% 
  group_by(continent) %>% 
  count(n_distinct(country))
```


*Can I visualize this in a graph?*
Note: divide by 12 because there are 12 time points per country
```{r}
ggplot(gapminder) + 
  geom_bar(aes(continent, ..count../12, fill=continent)) +
  xlab("Continent") +
  ylab("Number of Countries") +
  theme_bw()

```


#### GDP per capita

*Let's get run summary to find the range, median and mean for GDP per capita*
```{r}
gapminder %>% 
  select(gdpPercap) %>% 
  summary() 
```

*Which country had the lowest GDP per capita and which country had the highest GDP per capita in this dataset?*
```{r}
gapminder %>% 
  select(country, year, gdpPercap) %>% 
  filter(gdpPercap == min(gdpPercap) | gdpPercap == max(gdpPercap)) %>% 
    group_by(country) 
```
The Democratic Republic of Congo (in 2002) had the lowest GDP per capita. Kuwait (in 1957) had the highest GDP per capita.


*What other countries had the lowest GDP per capita?*
```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(min_GDP = min(gdpPercap),
            country = country[gdpPercap == min_GDP]) %>% 
  arrange(min_GDP)
```


*We can also see when each country had their lowest verses their highest GDP per capita*
```{r}
gapminder %>% 
  select(country, year, gdpPercap) %>% 
  group_by(country) %>% 
  filter(gdpPercap == min(gdpPercap) | gdpPercap == max(gdpPercap)) 
```



# Exercise #3: Explore various plot types

####  GDP by life expectancy in 2007
```{r}
gapminder %>% 
  filter(year == 2007) %>% 
  ggplot(aes(gdpPercap,lifeExp)) +
  geom_point(aes(color=continent)) +
  theme_bw() +
  labs(x="GDP Per Capita", y="Life Expectancy")
```


#### Median GDP per capita by continent over time 

```{r}
gapminder %>% 
  group_by(year, continent) %>% 
  summarize(median_GDP = median(gdpPercap)) %>% 
  ggplot(aes(x=year, y=median_GDP, colour=continent)) +
  geom_line() +
  theme_bw() +
  labs(x="Year", y ="Median GDP Per Capita") 
```


