---
title: "Homework 02"
author: "Kimberly Sharpe"
output:
  html_document:
    keep_md: yes
    theme: paper
  output:
    html_document:
     df_print: paged
---

<!--- allow for errors while knitting-->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE)
```

<!--- load packages needed for the assignment ---> 

```{r load packages, echo=FALSE, message=FALSE}
library(gapminder)
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(scales))
```
 
# Exercise 1: Basic dplyr 

## 1.1 subsetting gapminder to three countries in the 1970s
```{r choose 3 countries in 1970s}
gapminder %>% 
  filter(year > 1969 & year < 1980, 
        country == "Afghanistan" |
        country == "Canada" | 
        country == "Denmark")
```
## 1.2 using a pipe operator to select country & gdpPercap from filtered dataset
 
```{r}
gapminder %>% 
  filter(year > 1969 & year < 1980, 
        country == "Afghanistan" |
        country == "Canada" | 
        country == "Denmark") %>% 
  select(country, gdpPercap)
  
```

## 1.3 filtering gapminder to all entries that have experienced a drop in life expectancy


```{r}
gapminder %>% 
  select(country, year, lifeExp) %>% 
  arrange(country, year) %>% 
  group_by(country) %>% 
  mutate(diff_LifeExp = lifeExp - lag(lifeExp)) %>% 
  filter(diff_LifeExp < 0)
  
```


## 1.4 showing max GDP per capita experienced by each country 
```{r}
gapminder %>% 
  group_by(country) %>% 
  summarise(max_gdpPercap = max(gdpPercap)) # %>%
 # arrange(max_gdpPercap)  ##just for fun
```

## 1.5 producing a scatterplot of Canada's life expectancy vs GDP 
Ensure log scale
Pipe directly into ggplot after altering dataset

```{r}
gapminder %>% 
  filter(country == "Canada") %>% 
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point(color = "blue") +
  scale_x_log10("GDP per capita", labels = scales::dollar_format()) +
  theme_bw() +
  ylab("Life Expectancy")
```


# Exercise 2: Explore individual variables with dplyr

**Choose one categorial and one quantitative variable:**
Categorical variable: continent
Quantative variable: gdpPercap

## What are the possible values of each variable?

**Continent:** 

How many continents are in the dataset?
```{r}
gapminder %>% 
  select(continent) %>% 
  summarize(n_unique = n_distinct(continent))

```

What continents are included in the dataset?
```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarize(n_unique = n_distinct(continent))

```

How many countries are in each continent?

```{r}
gapminder %>% 
  select(continent, country) %>% 
  group_by(continent) %>% 
  summarize(n_unique = n_distinct(country))
```

Can I visualize this in a graph?
Note: divide by 12 because there are 12 time points per country

```{r}
ggplot(gapminder, aes(fill=continent)) + 
  geom_bar(aes(continent, ..count../12)) +
  xlab("Continent") +
  ylab("Number of Countries")

```


**GDP per capita**

```{r}
gapminder %>% 
  select(gdpPercap) %>% 
  range()
```

```{r}
gapminder %>% 
  select(gdpPercap) %>% 
  median() %>% 
```


```{r}
ggplot(data=gapminder, aes(x=gdpPercap, y=year)) +
    geom_boxplot(alpha=0.3)
```


What values are typical? What's the spread? What's the distribution? (tailored to each variable at hand)


# Exercise #3: Explore various plot types

##  life expectancies of continents over time?

## population of each continent over time?