---
title: "HW 03"
output:
  html_document:
    keep_md: yes
    theme: paper
  pdf_document: default
---

<!---The following chunk allows errors when knitting--->

```{r allow errors, echo = FALSE}
knitr::opts_chunk$set(error = TRUE, warning = FALSE)
```

<!-- load the required packages! --->
```{r, message = FALSE, warning = FALSE}
library(gapminder)
library(tibble)
library(tidyverse)
library(ggplot2)
library(scales)
```

# Assignment #3 - Choose three coding options 

##  Task Option 2: Get the maximum and minimum GDP per capita for all continents

I chose this option because for a previous assignment I looked at the minimum and maximum GDP per capita overall. Now I'd like to figure out how to do it for each continent. 

<!---
So this code works, but it's not pretty... is there another way?
```{r}
gapminder %>% 
  select(continent, year, gdpPercap) %>% 
  group_by(continent) %>% 
  filter(gdpPercap == min(gdpPercap) | gdpPercap == max(gdpPercap))
```
--->

<!--- The summarize function makes more sense to use although it doesn't display the year for the min and max GDP, which would be interesting to know. 
Also don't need to select() in this iteration. --->

**First, calculate the minimum and maximum GDP per capita using the summarize() function.**
```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarize(min_GDP = min(gdpPercap), max_GDP = max(gdpPercap))
```


**Second, visualizing this data in a graph:**

<!--- Note to self: the alpha needs to be outside the aes() function or else it'll show up as a legend on the side.
https://stackoverflow.com/questions/11714951/remove-extra-legends-in-ggplot2 --->


<!--- https://stackoverflow.com/questions/33221794/separate-palettes-for-facets-in-ggplot-facet-grid/33222028#33222028 <-- It looks like it's difficult to get a good looking colour gradient. Maybe next time... --->

<!--- learned through accidentally keeping both the min and max geom_col coding lines that I can use that to plot both ---> 

```{r}
gapminder %>% 
  group_by(continent) %>% 
  summarize(min_GDP = min(gdpPercap), max_GDP = max(gdpPercap)) %>% 

ggplot(aes(colour=GDP)) +
  geom_col(aes(continent, min_GDP, colour = "Min GDP"), alpha = 0.4) +
  geom_col(aes(continent, max_GDP, colour = "Max GDP"), alpha = 0.4) +
  theme_bw() +
  scale_y_log10(labels = scales::dollar_format()) +
  labs(x="Continent", y="(log) GDP Per Capita") +
  ggtitle("Minimum and Maximum GDP Per Capita for Each Continent")
```

**Finally, some thoughts about what the data says:**
Asia has the biggest discrepnacy between minimum GDP and maximum GDP per capita. 


# Task 5: How is life expectancy changing over time on different continents?

To examine this, I am looking at the median life expectancy.

**First, using the group_by() and summarize() functions to look at life expectancy over time**
```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(lifeExp = median(lifeExp)) %>% 
  DT::datatable()
```

**Second, let's visualize the data.**
<!--- for future reference on how to put labels at the end of plot points:
https://stackoverflow.com/questions/29357612/plot-labels-at-ends-of-lines ---> 

```{r}
gapminder %>% 
  group_by(year, continent) %>% 
  summarize(lifeExp = median(lifeExp)) %>% 
  ggplot(aes(x=year, y=lifeExp, colour=continent)) +
  geom_line() +
  theme_bw() +
#  geom_dl(aes(label = continent), method = list(dl.trans(x = x + .2), "last.points")) + couldn't figure out how to get the labels within the graph confines
#  facet_wrap(~ continent, scales = "free_y") + liked it better all on one graph for comparison
#  theme(legend.position = "none") +
  labs(x="Year", y ="Median Life Expectency") +
  ggtitle("Life Expectancy Over Time")
```

The table and graph suggest that for ...




# Task 4: Compute a trimmed mean of life expectancy for different years 
<!--- https://www.rdocumentation.org/packages/DescTools/versions/0.99.19/topics/Trim how to use trim() function ---> 

**First, calculate trimmed means across all continents for each year using trim in the mean() function.**
```{r}
gapminder %>% 
  group_by(year) %>% 
  summarize(trim_LifeExp = mean(lifeExp, trim = 0.10)) 
# summarize(trim_LifeExp = mean(lifeExp, trim =3)) can also trim a specific number from each end.
```


**Second, plot the the trimmed means.**

```{r}
gapminder %>% 
  group_by(year) %>% 
  summarize(trim_LifeExp = mean(lifeExp, trim = 0.10)) %>% 

  ggplot(aes(x=factor(year), y=trim_LifeExp)) +
  geom_point(colour = "darkseagreen4", size = 3, shape = 21, stroke = 2) +
  labs(x = "Year", y = "Mean (trimmed)") +
  theme(legend.position = "none") +
  theme_bw() +
  ggtitle("Mean Life Expectancy Over Time for All Continents")


```

<!---
Plotting trimmed means by year - alternative: 
I came across an alternative on the old STAT 545 message board. https://github.com/STAT545-UBC/Discussion/issues/132 & https://ggplot2.tidyverse.org/reference/stat_summary.html and decided this may be a tidier way to display the data --->






