---
title: "hw01_gapminder"
output: 
  html_document:
    keep md: true
    theme: cerulean
---


<!-- load packages and gapminder dataset  -->

```{r load packages, echo=FALSE}
library(tibble)
library(DT)
library(gapminder)
library(ggplot2)
```

## For hw01, we are going to be exploring the gapminder dataset



### First, let's convert gapminder to a tibble and wrap it in a datatable

```{r}
datatable(as_tibble(gapminder))
```


### Next, let's explore the structure of the dataset


What are the column names in our dataset?

```{r}
names(gapminder)
```


What types of variables does the dataset contain?

```{r}
str(gapminder)
```


Let's get a summary of the the dataset

```{r}
summary(gapminder)
```



## Next, let's do some basic analyses


How many countries are in our dataset?

```{r}
length(unique(gapminder$country))
```


What is the range of GDP in our dataset? and life expectancy in our dataset?

```{r}
range(gapminder$gdpPercap)
```


What about the range for life expectacy in our dataset?

```{r}
range(gapminder$lifeExp)
```



Finally, let's look at life expectancy by continent
```{r}
ggplot(gapminder, aes(continent, lifeExp, fill=continent)) +
  geom_boxplot() +
  theme_bw() +
  xlab("Continent") +
  ylab("Life Expectancy (years)")
```


